---
title: "Introduction to R & RStudio"
author: "Steve Pederson"
date: "1 May, 2017"
output: 
  ioslides_presentation: 
    css: default.css
    fig_caption: yes
    keep_md: yes
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, 
											eval = TRUE, results = "markup",
											message = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
x <- 2
y <- 3
```


# Vectors

## The R Environment | `c()`

* One of the most useful functions is `c()`
    + Stands for `combine` or `concatenate`
    + We can join two (or more) objects into one object
    
```{r}
z <- c(x, y)
z
```

## Vectors

* The ability for an `R` object to hold more than one value:
    + Vectors are the fundamental unit in `R`
    + Are a 1-dimensional set of values
    + We can apply functions to an entire vector

```{r, eval=FALSE}
sqrt(z)
2^z
length(z)
```

## Vectors

### What issues might crop up when applying a function to a vector?

Hint: What types of values exist?

## Vectors

### What issues might crop up when applying a function to a vector?


* Can we perform mathematical calculations on words?
* Can we use numbers with decimal points to indicate positions in a ranked list?

## Vectors | 6 atomic vectors

`R` has 6 atomic vector types

*  

