---
title: "Subsetting, dplyr, magrittr"
author: "Lloyd Low"
output:
  pdf_document: default
  html_document:
    keep_md: yes
  word_document: default
---

# Introduction
So you have got a table with data that might be a mixed of categorical, integer, numeric, etc variables? And it comes with thousands of columns and millions of rows? You wish you can manipulate your table in 'powerful' ways beyond just the simple preconfigured operations that some software has already set for you, which does not allow you to ask questions in a different way? Below is a potential solution using R, which is open source and that means it's 'almost' free as long as you adhere to the license conditions.

*Interesting fact!* Did you know what Microsoft Excel limits are? 
1,048,576 rows by 16,384 columns

* Outline of topics
    + Logical Vectors
    + Subsetting using square brackets
    + dplyr - Who created it? Hadley Wickham and Romain Francois
    + dplyr - What is it? Grammar for manipulating tabular data
    + dplyr functions: select(), filter(), arrange(), mutate(), group_by(), summarise()
    + Pipe a series of dplyr functions using the magrittr (%>%) package

# Subsetting
```{r, echo = TRUE, fig.width=3, fig.height = 3, fig.align='center'}
#subset a vector#
x <- c("a", "b", "c", "d", "a", "b", "d")
x[1]
x[1:5]
x[x>"a"]
boolean <- x > "a"
boolean
x[boolean]

#removal of NAs
z <- c(1,2,NA,4,NA,6,7,8,9)
bad <- is.na(z)
z[!bad]

z <- c(1,2,NA,4,NA,6,7,8,9)
y <- c("a","b",NA,"d",NA,"f","g","h","i")
good <- complete.cases(z,y)
z[good]
y[good]

#subset a data frame#
setwd("/Users/lloyd/Documents/lloyd_2017/Services/R_RoseworthyWkshp_git/Roseworthy-R-20170501/data")

#read in the data frame
tooth <- read.csv("toothData.csv")

#Check for missing data
tooth.na <- is.na(tooth$len)
sum(tooth.na)

#Alternative way to subset
tooth.na <- is.na(tooth[,1])
sum(tooth.na)

#how to select just the columns 'len' and 'dose'?
tooth.len.dose <- tooth[,c(1,3)]

#how to filter for a particular row? say tooth length more than mean tooth length?
tooth.len.abv.mean <- tooth[tooth$len > mean(tooth$len),]
```
# dplyr
```{r, echo = TRUE, fig.width=3, fig.height = 3, fig.align='center'}
library(dplyr)
#library(data.table) #Use this package if the data frame is very big

#Convert it to a class that dplyr can work on
tooth <- tbl_df(tooth)
#observe that it is class tibble
class(tooth)

#five verbs (technically called functions in R)
#select(), filter(), arrange(), mutate(), and summarize()

select(tooth,len,dose)
#dplyr is 'clever' enough to display just the first 10
#rows of data and columns that can fit neatly in the console. 
#For missed columns due to limited console viewing space,
#we see the names and classes for the variables at the bottom.

#select() will output according to the order of variables we specify
select(tooth,dose,len)

#using ':' operator, can select a range of variables
select(tooth,len:dose)

#using '-' operator to throw away unwanted columns
select(tooth,-dose)

#filter specific rows
filter(tooth, len == 10)
?Comparison

#arrange len in ascending order
arrange(tooth, len)

#Suppose the tooth length is given in mm, and you want a len column in cm
mutate(tooth, len.cm = len / 10)

#group_by and summarise
tooth.supp <- group_by(tooth,supp,dose)
summarise(tooth.supp,mean(len))
```

# Exercise
```{r, echo = TRUE, fig.width=9, fig.height = 6, fig.align='center', cache=TRUE}
#Exercise
#read in ChickWeightNew.csv and make the data okay for dplyr to work with

#select just weight and Chick columns from the dataframe/tibble

#remove observations that have weight equals to "NA"

#remove observations that have weight equals to "NA" and keep column "Diet" equals to 1

#filter rows with Time = 21 Or Diet = 2

#arrange the ChickWeight dataframe according to Time in ascending order

#arrange the ChickWeight dataframe according to Time in ascending order and 
#Diet in descending order

#filter Time = 10, save dataframe as a new one, 
#create a new variable that is the normalized weight, 
#i.e. (weight - mean(weight))/sd(weight)

#remove all NA from weight, filter for last day = 21, group by diet, summarize mean diet 

```

# Answers
```{r, echo = TRUE, fig.width=9, fig.height = 6, fig.align='center', cache=TRUE}
#read in ChickWeight
setwd("/Users/lloyd/Documents/lloyd_2017/Services/R_RoseworthyWkshp_git/Roseworthy-R-20170501/data")
test <- read.csv("ChickWeightNew.csv")
library(dplyr)
test <- tbl_df(test)

#select just weight and Chick
select(test,weight, Chick)

#remove rows with "NA" for weight
filter(test, weight != "NA")
#Alternative way
filter(test, !is.na(weight))

#remove rows that don't have "NA" and the "Diet" is equal to 1
#combine filter based on AND logic
filter(test, weight != "NA",Diet == 1)

#Time = 21 Or Diet = 2
#use OR logic |
filter(test, Time == 21 | Diet == 2)

#arrange the DF according to Time in ascending order
arrange(test, Time)

#can do multiple variable arrangement
#arrange the DF according to Time in ascending order and Diet in descending order
arrange(test, Time,desc(Diet))

#mutate - create new variable based on existing ones
#filter only Time = 10, save DF, create a new variable that is the normalized
#weight, i.e. (weight - mean(weight))/sd(weight)
test.10 <- filter(test,Time == 10)
mutate(test.10,normalizedWeight = (weight - mean(weight))/sd(weight))

#group_by and summarize
#remove NA, filter for last day = 21, group by diet, summarize mean diet 
test <- filter(test, !is.na(weight))
test.lastday <- filter(test,Time == 21)
test.lastday.diet <- group_by(test.lastday,Diet)
summarise(test.lastday.diet,mean(weight))

```

# magrittr %>%
```{r, echo = TRUE, fig.width=9, fig.height = 6, fig.align='center', cache=TRUE}
setwd("/Users/lloyd/Documents/lloyd_2017/Services/R_RoseworthyWkshp_git/Roseworthy-R-20170501/data")
test <- read.csv("ChickWeightNew.csv")
library(dplyr)
library(magrittr)
test <- tbl_df(test)

#filter only Time = 10, save DF, create a new variable that is the normalized
#weight, i.e. (weight - mean(weight))/sd(weight)
test %>% filter(Time == 10) %>% mutate(normalizedWeight = (weight - mean(weight))/sd(weight))
#group_by and summarize
#remove NA, filter for last day = 21, group by diet, summarize mean diet 
test %>% filter(!is.na(weight),Time == 21) %>% group_by(Diet) %>% summarise(mean(weight))
```